Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-04-23 manuscript.qmd
finalize uni model recap

$$
\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
$$

@@

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. This effectively masks large parts of the transition matrix to be zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

@@

@@

Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

The quantity $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, our parameter space is:

$$
\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, \left\{p_\text{T}x \right\} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
$$

@@

$$
\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
$$

@@

<!-- TODO: Explain here what role our model plays in a Bayesian context. -->

$$
P \left( \mathbf{X} \mid \mathbf{Z} \right) = \frac{P \left( \mathbf{Z} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{P \left( \mathbf{Z} \right)}
$$ {#eq-uni-bayes-law}

Let us for later use define the matrices $\boldsymbol{\Lambda}$:

The $n$-th row in this matrix corresponds to the distribution over hidden states after $t=n-1$ time-steps.

@@

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
$$

>>>>

$$
\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
$$ {#eq-state-def}

@@

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-trans-matrix}
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. This effectively masks large parts of the transition matrix to be zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-obs-matrix}
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

@@

For later use, we define at this point the matrices $\boldsymbol{\Lambda}$:

Where the $k$-th row in this matrix corresponds to the distribution over hidden states after $t=k-1$ time-steps.

@@

Above, the quantity inside $\left[ \ldots \right]_i$ denotes the $i$-th component of the vector that is the result of the vector and matrix multiplications in the square brackets. Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, our parameter space is:

$$
\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
$$ {#eq-param-space}

@@

$$
\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
$$ {#eq-log-likelihood}

The methodology we use to infer the model's parameters is detailed in @sec-sampling.

@@

Our stated goal is to compute the risk for a patient's true nodal involvement state $\mathbf{X}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}_k, \text{T}x \right)$. Using Bayes' law, this is written as:

$$
P \big( \mathbf{X} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \frac{P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right) P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}{\sum_{i=0}^{2^V} P \left( \boldsymbol{\zeta}_k \mid \mathbf{X}=\boldsymbol{\xi}_i \right) P \big( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
$$ {#eq-uni-bayes-law}

The term $P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right)$ is defined solely by sensitivity and specificity of the diagnostic modality. Terms like this already appeared in the definition of the observation matrx in @eq-obs-matrix. The *prior* $P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}} \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.

It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $\mathbf{X}$, but also for e.g. individual LNLs: For example, the risk for involvement in level IV would be a marginalization over all states $\boldsymbol{\xi}_i$, where $\xi_{i4}=1$. Formally:

$$
P \big( \text{IV} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big) = \sum_{k \, : \, \xi_{k4}=1} P \big( \mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big)
$$ {#eq-marg-over-posterior}

@@

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
$$ {#eq-bilateral-bayes}

For the sake of brevity, we omit the dependency on the parameters and the T-category here.
